# %% [code]
---
title: "Cyclistic Bikeshare"  
subtitle: Capstone
author: "T. Green"

output:
	html_document:
		number_sections: true
		toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Google Profressional Data Analytics Capstone
## Environment setup

```{r chunk-1}

## ------------------------------------------------------
## Load required libraries
## Suppress warnings
## ------------------------------------------------------


library(tidyverse)
library(rmarkdown)


options(warn = -1)

```



```{r}

## ------------------------------------------------------
## List files
## ------------------------------------------------------


list.files(path = "../input/cyclistic-capstone")


```



## Examine Flat Files

```{r chunk-3}

## ------------------------------------------------------
## Assign csv file to y2021_05 for a preliminary view 
## View first six rows of data
## View file structure
## View summary of data
## Imported 531,633 rows
## ------------------------------------------------------


y2021_05 <- read_csv(file = "../input/cyclistic-capstone/202105-divvy-tripdata.csv")
head(y2021_05)
str(y2021_05)
summary(y2021_05)

```


```{r chunk-4}

## ------------------------------------------------------
## Assign csv file to y2021_06 for a preliminary view 
## View first six rows of data
## View file structure
## View summary of data
## Imported 729,595 rows
## ------------------------------------------------------


y2021_06 <- read_csv(file = "../input/cyclistic-capstone/202106-divvy-tripdata.csv")
head(y2021_06)
str(y2021_06)
summary(y2021_06)


```



```{r chunk-5}

## ------------------------------------------------------
## Assign csv file to y2021_07 for a preliminary view 
## View first six rows of data
## View file structure
## View summary of data
## Imported 822,410 rows
## ------------------------------------------------------


y2021_07 <- read_csv(file = "../input/cyclistic-capstone/202107-divvy-tripdata.csv")
head(y2021_07)
str(y2021_07)
summary(y2021_07)

```

```{r chunk-6}

## ------------------------------------------------------
## Assign csv file to y2021_08 for a preliminary view 
## View first six rows of data
## View file structure
## View summary of data
## Imported 804,352 rows
## ------------------------------------------------------


y2021_08 <- read_csv(file = "../input/cyclistic-capstone/202108-divvy-tripdata.csv")
head(y2021_08)
str(y2021_08)
summary(y2021_08)

```


```{r chunk-7}

## ------------------------------------------------------
## Assign csv file to y2021_09 for a preliminary view 
## View first six rows of data
## View file structure
## View summary of data
## Imported 756,147 rows
## ------------------------------------------------------


y2021_09 <- read_csv(file = "../input/cyclistic-capstone/202109-divvy-tripdata.csv")
head(y2021_09)
str(y2021_09)
summary(y2021_09)

```


```{r chunk-8}

## ------------------------------------------------------
## Assign csv file to y2021_10 for a preliminary view 
## View first six rows of data
## View file structure
## View summary of data
## Imported 631,226 rows
## ------------------------------------------------------


y2021_10 <- read_csv(file = "../input/cyclistic-capstone/202110-divvy-tripdata.csv")
head(y2021_10)
str(y2021_10)
summary(y2021_10)

```


```{r chunk-9}

## ------------------------------------------------------
## Assign csv file to y2021_11 for a preliminary view 
## View first six rows of data
## View file structure
## View summary of data
## Imported 359,978 rows
## ------------------------------------------------------


y2021_11 <- read_csv(file = "../input/cyclistic-capstone/202111-divvy-tripdata.csv")
head(y2021_11)
str(y2021_11)
summary(y2021_11)

```


```{r chunk-10}

## ------------------------------------------------------
## Assign csv file to y2021_12 for a preliminary view 
## View first six rows of data
## View file structure
## View summary of data
## Imported 247,540 rows
## ------------------------------------------------------


y2021_12 <- read_csv(file = "../input/cyclistic-capstone/202112-divvy-tripdata.csv")
head(y2021_12)
str(y2021_12)
summary(y2021_12)


```


```{r chunk-11}

## ------------------------------------------------------
## Assign csv file to y2022_01 for a preliminary view 
## View first six rows of data
## View file structure
## View summary of data
## Imported 103,770 rows
## ------------------------------------------------------


y2022_01 <- read_csv(file = "../input/cyclistic-capstone/202201-divvy-tripdata.csv")
head(y2022_01)
str(y2022_01)
summary(y2022_01)


```


```{r chunk-12}

## ------------------------------------------------------
## Assign csv file to y2022_02 for a preliminary view 
## View first six rows of data
## View file structure
## View summary of data
## Imported 115,609 rows
## ------------------------------------------------------


y2022_02 <- read_csv(file = "../input/cyclistic-capstone/202202-divvy-tripdata.csv")
head(y2022_02)
str(y2022_02)
summary(y2022_02)


```


```{r chunk-13}

## ------------------------------------------------------
## Assign csv file to y2022_03 for a preliminary view 
## View first six rows of data
## View file structure
## View summary of data
## Imported 284,042 rows
## ------------------------------------------------------


y2022_03 <- read_csv(file = "../input/cyclistic-capstone/202203-divvy-tripdata.csv")
head(y2022_03)
str(y2022_03)
summary(y2022_03)

```

```{r chunk-14}

## ------------------------------------------------------
## Assign csv file to y2022_04 for a preliminary view 
## View first six rows of data
## View file structure
## View summary of data
## Imported 371,249 rows
## ------------------------------------------------------


y2022_04 <- read_csv(file = "../input/cyclistic-capstone/202204-divvy-tripdata.csv")
head(y2022_04)
str(y2022_04)
summary(y2022_04)

```

## Merge Flat Files


```{r  chunk-15}

## -------------------------------------------------------
## Merge ALL csv files 
## Imported 5,757,551 records
## -------------------------------------------------------

cyclistic_merged <- rbind(y2021_05
                          , y2021_06
                          , y2021_07
                          , y2021_08
                          , y2021_09
                          , y2021_10
                          , y2021_11
                          , y2021_12
                          , y2022_01
                          , y2022_02
                          , y2022_03
                          , y2022_04
)

```


## Data Transformation


Load additional libraries required for data transformation.




```{r   chunk-16}

## ------------------------------------------------------
## Load required libraries
## Suppress warnings
## ------------------------------------------------------

library(here)
library(skimr)
library(dplyr)
library(janitor)
library(data.table)
library(geosphere)

```


```{r  chunk-17}

## -------------------------------------------------------
## Remove non character spaces from data set
## -------------------------------------------------------

clean_names(cyclistic_merged)

```


```{r   chunk-18}

## -------------------------------------------------------
## Inspect unique values in variable
## Rideable type | start station name | start station id
## End station name | end station id | member casual
## - Divvy, Chargingstx, Temp, Corporate Private racks and
## NA fields - ** remove after updates are made **
## -------------------------------------------------------


unique(cyclistic_merged$rideable_type)
unique(cyclistic_merged$member_casual)


#### omitted for notebook ####

# unique(cyclistic_merged$start_station_name) 
# unique(cyclistic_merged$start_station_id)
# unique(cyclistic_merged$end_station_name)
# unique(cyclistic_merged$end_station_id)


```


```{r   chunk-19}

## -------------------------------------------------------
## Create a copy of start stations to be removed 
## Returned 29,918 records 
## - Divvy, Chargingstx, Temp, and Corporate Private racks
## - NA station information to be checked last
## -------------------------------------------------------


start_stations <- cyclistic_merged %>% 
  filter(start_station_id == "Hubbard Bike-checking (LBS-WH-TEST)" |
           start_station_id == "Bissell St & Armitage Ave - Charging" |
           start_station_id == "charginstx1" |
           start_station_id == "DIVVY CASSETTE REPAIR MOBILE STATION" |
           start_station_id == "charginstx3" |
           start_station_id == "Hastings WH2" |
           start_station_id == "Lincoln Ave & Rosco St - Charging" |
           start_station_id == "chargingstx5" |
           start_station_id == "chargingstx4" |
           start_station_id == "Pawel Bialowas - Test- PBSC charging station" |
           start_station_id == "DIVVY 001" |
           start_station_id == "Wilton Ave & Diversey Pkwy - Charging" |
           start_station_id == "13259" |
           start_station_id == "TA1309000008" |
           start_station_id == "13099" |
           start_station_id == "KA1504000141" |
           start_station_id == "TA1308000026" |
           start_station_id == "13285" |
           start_station_id == "20999"
)


```




```{r chunk-20}

## -------------------------------------------------------
## Remove 29,918 records 
## Divvy, Chargingstx, Temp, and Corporate Private racks
## Remaining - 5,727,633 records
## -------------------------------------------------------


cyclistic_merged <- cyclistic_merged %>% anti_join(start_stations)


```


```{r chunk-21}

## -------------------------------------------------------
## Create a copy of end stations to be removed 
## Returned 28,200 records 
## - Divvy, Chargingstx, Temp, and Corporate Private racks
## -------------------------------------------------------


end_stations <- cyclistic_merged %>% 
  filter(end_station_id == "Hubbard Bike-checking (LBS-WH-TEST)" |
           end_station_id == "Bissell St & Armitage Ave - Charging" |
           end_station_id == "charginstx1" |
           end_station_id == "DIVVY CASSETTE REPAIR MOBILE STATION" |
           end_station_id == "charginstx3" |
           end_station_id == "Hastings WH2" |
           end_station_id == "Lincoln Ave & Rosco St - Charging" |
           end_station_id == "chargingstx5" |
           end_station_id == "chargingstx4" |
           end_station_id == "Pawel Bialowas - Test- PBSC charging station" |
           end_station_id == "DIVVY 001" |
           end_station_id == "Wilton Ave & Diversey Pkwy - Charging" |
           end_station_id == "13259" |
           end_station_id == "TA1309000008" |
           end_station_id == "13099" |
           end_station_id == "KA1504000141" |
           end_station_id == "TA1308000026" |
           end_station_id == "13285" |
           end_station_id == "20999"
  )

```


```{r chunk-22}

## -------------------------------------------------------
## Remove 28,200 records 
## Divvy, Chargingstx, Temp, and Corporate Private racks
## Remaining - 5,699,433 records
## -------------------------------------------------------


cyclistic_merged <- cyclistic_merged %>% anti_join(end_stations)


```


```{r chunk-23}


## -------------------------------------------------------
## Search for missing start station names with station id 
## Imported 3 records
## Station id: 13221 |  20215   | WL-008
## -------------------------------------------------------


cyclistic_merged %>%
  filter(is.na(start_station_name) & !is.na(start_station_id))  

```


```{r chunk-24}

## -------------------------------------------------------
## Search for station id 13221 
## Start station name = Wood St & Milwaukee Ave
## -------------------------------------------------------


head(cyclistic_merged %>%
       filter(start_station_id == 13221) %>% 
       select(start_station_name), 10)


```


```{r chunk-25}

## -------------------------------------------------------
## Search for station id 20215 
## Start station name = Hegewisch Metra Station
## -------------------------------------------------------


head(cyclistic_merged %>%
       filter(start_station_id == 20215) %>% 
       select(start_station_name), 10)


```


```{r chunk-26}

## -------------------------------------------------------
## Search for station id WL-0008 
## Start station name = Clinton St & Roosevelt Rd
## -------------------------------------------------------

head(cyclistic_merged %>%
       filter(start_station_id== "WL-008") %>% 
       select(start_station_name), 10)


```


```{r chunk-27}

## -------------------------------------------------------
## Update missing station names
## Cleared 788,155 start station NA values
## Cleared 840,793 end station NA values
## Skim for updates
## -------------------------------------------------------


cyclistic_merged$start_station_name[ 
  which(is.na(
    cyclistic_merged$start_station_id == 13221))] <- "Wood St & Milwaukee Ave"   


cyclistic_merged$end_station_name[ 
  which(is.na(
    cyclistic_merged$end_station_id == 13221))] <- "Wood St & Milwaukee Ave"   


skim_without_charts(cyclistic_merged)


```


```{r chunk-28}

## -------------------------------------------------------
## Inspect remaining null start station names
## 176105D1F8A1216B record  - station id 13221
## DE82A15026BA3056 record  - station id 20215
## EE197EDA4CF8CFE5 record  - station id WL-008
## -------------------------------------------------------


cyclistic_merged %>% 
        select(start_station_name, start_station_id, start_lat, start_lat, everything()) %>% 
        filter(is.na(start_station_name))

```


```{r chunk-29}

## -------------------------------------------------------
## Update remaining 3 start station names
## 176105D1F8A1216B record - Wood St & Milwaukee Ave | 13221
## DE82A15026BA3056 record - Hegewisch Metra Station | 20215
## EE197EDA4CF8CFE5 record - Clinton St & Roosevelt Rd | WL-008
## -------------------------------------------------------


cyclistic_merged$start_station_name <- 
  ifelse(cyclistic_merged$ride_id == "176105D1F8A1216B", 
         "Wood St & Milwaukee Ave", 
         cyclistic_merged$start_station_name)


cyclistic_merged$start_station_name <- 
  ifelse(cyclistic_merged$ride_id == "DE82A15026BA3056", 
         "Clinton St & Roosevelt Rd", 
         cyclistic_merged$start_station_name)


cyclistic_merged$start_station_name <- 
  ifelse(cyclistic_merged$ride_id == "EE197EDA4CF8CFE5", 
         "Wood St & Milwaukee Ave", 
         cyclistic_merged$start_station_name)


skim_without_charts(cyclistic_merged)

```


```{r chunk-30}

## -------------------------------------------------------
## Inspect NA start station id
## Start station name = Wood St & Milwaukee Ave  | id = 13221
## Cleared 788,155 NA values
## Skim for updates
## -------------------------------------------------------


cyclistic_merged %>% 
  select(start_station_id, start_station_name, start_lat, start_lng) %>% 
  filter(is.na(start_station_id)) %>%
  group_by(start_station_name)


cyclistic_merged<-cyclistic_merged %>% 
  mutate(start_station_id = ifelse(start_station_name == "Wood St & Milwaukee Ave", 
                                   "13221", start_station_id))


skim_without_charts(cyclistic_merged) # skim after changes to ensure desired change occured


```


```{r chunk-31}

## -------------------------------------------------------
## Inspect NA end station id
## End station name = Wood St & Milwaukee Ave  | id = 13221
## Cleared 840,793 NA values
## Skim for updates
## -------------------------------------------------------


cyclistic_merged %>% 
  select(end_station_id, end_station_name, end_lat, end_lng) %>% 
  filter(is.na(end_station_id)) %>%
  group_by(end_station_name)


cyclistic_merged<-cyclistic_merged %>% 
  mutate(end_station_id = ifelse(end_station_name == "Wood St & Milwaukee Ave", 
                                 "13221", end_station_id))


skim_without_charts(cyclistic_merged)


```


```{r chunk-32}

## -------------------------------------------------------
## Inspect NA end latitude and longitude
## End latitude and longitude = Wood St & Milwaukee Ave 
## End lat = 41.9  |  end lng = -87.6
## Cleared 4,741 NA values for both end lat and end lng
## Skim for updates
## -------------------------------------------------------


cyclistic_merged %>% 
  select(end_station_id, end_station_name, end_lat, end_lng) %>% 
  filter(is.na(end_lat)) %>%
  group_by(end_station_name)


cyclistic_merged<-cyclistic_merged %>% 
  mutate(end_lat = ifelse(end_station_name == "Wood St & Milwaukee Ave", 41.9, end_lat)) %>% 
  mutate(end_lng = ifelse(end_station_name == "Wood St & Milwaukee Ave", -87.6, end_lng))


skim_without_charts(cyclistic_merged)

```


```{r chunk-33}

## -------------------------------------------------------
## Insert columns for analysis
## Date 
## Day of week
## Day of month day of the month 1-31 - space before single digits
## Month
## Hour
## Ride length
## Ride distance in miles
## Glimpse records
## Skim new columns
## -------------------------------------------------------


cyclistic_merged$date <- format(as.Date(cyclistic_merged$started_at), "%m-%d-%Y" )

cyclistic_merged$day_of_week <- format(as.Date(cyclistic_merged$started_at), "%A")

cyclistic_merged$day_of_month <- format(as.Date(cyclistic_merged$started_at), "%e") 

cyclistic_merged$month <- format(as.Date(cyclistic_merged$started_at), "%b '%y")

cyclistic_merged$hour <- as.numeric(hour(cyclistic_merged$started_at))

cyclistic_merged$ride_length <- as.numeric(difftime(cyclistic_merged$ended_at, 
                                                    cyclistic_merged$started_at, units = "mins"))


cyclistic_merged$ride_miles <- distGeo(matrix(c(cyclistic_merged$start_lng, 
                                                cyclistic_merged$start_lat), ncol = 2), 
                                       matrix(c(cyclistic_merged$end_lng, 
                                                cyclistic_merged$end_lat), 
                                              ncol = 2)) / 1609.35


cyclistic_merged <- cyclistic_merged %>% 
  mutate(round_trip=case_when(
    start_station_name==end_station_name ~ "roundtrip",
    start_station_id !=end_station_name ~ "oneway"))


glimpse(cyclistic_merged)
skim_without_charts(cyclistic_merged)


```


Remove Records outside of one (1) Day


```{r chunk-34}

## -------------------------------------------------------
## Create a copy of ride lengths under 1 minute
## Returned 91,109 records
## -------------------------------------------------------

under_min <- cyclistic_merged %>% 
  filter(ride_length <1)


```


```{r chunk-35}

## -------------------------------------------------------
## Remove 91,109 records 
## Rides less than 1 minute
## Remaining - 5,608,324 records
## -------------------------------------------------------


cyclistic_merged <- cyclistic_merged %>% anti_join(under_min)


skim_without_charts(cyclistic_merged)


```


```{r chunk-36}

## -------------------------------------------------------
## Create a copy of ride lengths over 1440 minutes (24 hrs)
## Returned 4,035 records
## -------------------------------------------------------


over_day <- cyclistic_merged %>% 
  filter(ride_length >1440)


```


```{r chunk-37}

## -------------------------------------------------------
## Remove 4,035 records 
## Rides greater than 1440 minutes (24 hrs)
## Remaining - 5,604,289 records
## -------------------------------------------------------


cyclistic_merged <- cyclistic_merged %>% anti_join(over_day)


skim_without_charts(cyclistic_merged)

```


```{r chunk-38}

## -------------------------------------------------------
## Change member casual variable types to factor and order
## member casual | rideable type | round trip | day of week
## -------------------------------------------------------


cyclistic_merged$member_casual <- ordered(
  cyclistic_merged$member_casual, levels = c("member", "casual"))


cyclistic_merged$rideable_type <- ordered(
  cyclistic_merged$rideable_type, levels = c("classic_bike", "electric_bike", 
                                             "docked_bike"))


cyclistic_merged$round_trip <- ordered(
  cyclistic_merged$round_trip, levels = c("oneway", "roundtrip"))



cyclistic_merged$day_of_week <- ordered(cyclistic_merged$day_of_week, levels = 
                                          c("Monday", "Tuesday", "Wednesday", 
                                            "Thursday", "Friday", "Saturday", "Sunday"))


str(cyclistic_merged)


skim_without_charts(cyclistic_merged)


```

## Descriptive Analysis



Load additional libraries required for descriptive analysis.



```{r chunk-39}

## ------------------------------------------------------
## Load required libraries
## Suppress exponential notation
## ------------------------------------------------------



library(DescTools)


options(scipen = 10)

```


```{r chunk-40}

## ------------------------------------------------------
## Compact structure overview 
## ------------------------------------------------------

Abstract(cyclistic_merged)


```


```{r chunk-41}

## ------------------------------------------------------
## Explore data set 
## ------------------------------------------------------

mean_length <- mean(cyclistic_merged$ride_length)
print(mean_length)  # Time difference of 19.02767 mins




mean_miles <- mean(cyclistic_merged$ride_miles)
print(mean_miles) # 1.973248


median_length <- median(cyclistic_merged$ride_length)
print(median_length)  # Time difference of 11.7 mins

median_miles <- median(cyclistic_merged$ride_miles)
print(median_miles) # 1.217151


Mode(cyclistic_merged$ride_length) # Mode = 5.783333 | freq: 5,698


Mode(cyclistic_merged$ride_miles)  # Mode = 0 | freq: 213,562


```


```{r chunk-42}

## ------------------------------------------------------
## Continue exploring data set 
## ------------------------------------------------------

cyclistic_merged %>% 
  tabyl(member_casual) 


```

```{r chunk-43}

cyclistic_merged %>% 
  tabyl(round_trip)

```


```{r chunk-44}

cyclistic_merged %>% 
  tabyl(round_trip, member_casual)

```


```{r chunk-45}

cyclistic_merged %>% 
  tabyl(rideable_type) 


```


```{r chunk-46}

cyclistic_merged %>% 
  tabyl(month) 


```


```{r chunk-47}

cyclistic_merged %>% 
  tabyl(day_of_week)


```


```{r chunk-48}

cyclistic_merged %>% 
  tabyl(day_of_month) 

```


```{r chunk-49}

cyclistic_merged %>% 
  tabyl(hour) 

```


```{r chunk-50}

## ------------------------------------------------------
## Order month for correct output. 
## Check order update
## ------------------------------------------------------


cyclistic_merged$month <- ordered(cyclistic_merged$month, levels = 
                      c("May '21", "Jun '21", "Jul '21", "Aug '21", "Sep '21", 
                        "Oct '21", "Nov '21", "Dec '21", "Jan '22", "Feb '22", 
                        "Mar '22", "Apr '22"))


cyclistic_merged %>% 
  tabyl(month) 



```


```{r chunk-51}

## ------------------------------------------------------
## Continue exploring data set
## ------------------------------------------------------

aggregate(cyclistic_merged$ride_length ~ cyclistic_merged$member_casual, FUN = mean)

```



```{r chunk-52}

aggregate(cyclistic_merged$ride_miles ~ cyclistic_merged$member_casual, FUN = mean)

```


```{r chunk-53}

aggregate(cyclistic_merged$ride_length ~ cyclistic_merged$member_casual + cyclistic_merged$day_of_week, FUN = mean)

```


```{r chunk-54}

aggregate(cyclistic_merged$ride_miles ~ cyclistic_merged$member_casual + cyclistic_merged$day_of_week, FUN = mean)

```



```{r chunk-55}

## ------------------------------------------------------
## Exploration continued
## ------------------------------------------------------

PercTable(x = cyclistic_merged$rideable_type, 
          y = cyclistic_merged$member_casual)


```


```{r chunk-56}

PercTable(x = cyclistic_merged$member_casual, 
          y = cyclistic_merged$round_trip)

```


```{r chunk-57}

PercTable(x = cyclistic_merged$rideable_type, 
          y = cyclistic_merged$round_trip)

```


```{r chunk-58}

PercTable(x = cyclistic_merged$member_casual, 
          y = cyclistic_merged$month)

```


```{r chunk-59}

PercTable(x = cyclistic_merged$member_casual, 
          y = cyclistic_merged$day_of_week)

```


```{r chunk-60}

PercTable(x = cyclistic_merged$member_casual, 
          y = cyclistic_merged$hour)

```


```{r chunk-61}

PercTable(x = cyclistic_merged$member_casual, 
          y = cyclistic_merged$day_of_month)


```



```{r chunk-62}

## ------------------------------------------------------
## Exploration wrap up
## ------------------------------------------------------

Desc(cyclistic_merged$rideable_type, main = "Rideable Type")


```


```{r chunk-63}

Desc(cyclistic_merged$rideable_type, main = "Rideable Type")


```


```{r chunk-64}

Desc(cyclistic_merged$member_casual, main = "Member vs. Casual Rider")



```


```{r chunk-65}

## ------------------------------------------------------
## Create member riders
## Imported 3,128,493 records 
## ------------------------------------------------------


member_riders <- cyclistic_merged %>% 
  filter(cyclistic_merged$member_casual == "member")


```



```{r chunk-66}

## ------------------------------------------------------
## Create casual riders
## Imported 2,475,796 records 
## ------------------------------------------------------


casual_riders <- cyclistic_merged %>% 
  filter(cyclistic_merged$member_casual == "casual")


```



```{r chunk-67}

Desc(cyclistic_merged$start_station_name, main = "Top Three Frequented Start Stations", head(3))

```

```{r chunk-68}


Desc(member_riders$start_station_name, main = "Top Three Member Frequented Start Stations", head(3))


```


```{r chunk-69}

Desc(casual_riders$start_station_name, main = "Top Three Casual Frequented Start Stations", head(3))

```

```{r chunk-70}

Desc(cyclistic_merged$end_station_name, main = "Top Three Frequented End Stations", head(3))

```


```{r chunk-71}

Desc(member_riders$end_station_name, main = "Top Three Member Frequented End Stations", head(3))

```


```{r chunk-72}


Desc(casual_riders$end_station_name, main = "Top Three Casual Frequented End Stations", head(3))


```



```{r chunk-73}

Desc(cyclistic_merged$month, main = "Rides by Month")


```

```{r chunk-74}

Desc(member_riders$month, main = "Member Rides by Month")

```


```{r chunk-75}

Desc(casual_riders$month, main = "Casual Rides by Month")


```

```{r chunk-76}

Desc(cyclistic_merged$ride_length, main = "Ride Length")

```


```{r chunk-77}

Desc(cyclistic_merged$ride_miles, main = "Ride Length in Miles")

```


```{r chunk-78}

Desc(cyclistic_merged$month, main = "Rides per Month")

```



```{r chunk-79}

Desc(cyclistic_merged$day_of_week, main = "Rides per Day")


```



## Visualize



Load additional libraries required for visualization of data.




```{r chunk-80}

## ------------------------------------------------------
## Load required libraries
## ------------------------------------------------------


library(ggplot2)
library(ggthemes)
library(scales)


```


```{r chunk-81}


cyclistic_merged %>% 
  group_by(member_casual, day_of_week) %>%
  summarise(num_rides = n(), .groups = 'drop') %>%
  ggplot(aes(x = day_of_week, y = num_rides, fill = member_casual,
             label = scales::comma(num_rides))) +
  geom_col(position = "dodge", width = .5) +
  geom_text(position = position_dodge(width = .9), # move to center of bars
            vjust = -0.5, # nudge above top of bar
            size = 3) +
  scale_y_continuous( labels = scales::comma) +
  labs(
    title = "Number of Rides",
    subtitle = "by weekday",
    caption = "Data source: Motivate International Inc. Figure 1",
    x = " ",
    y = "Number of Rides") +
theme(legend.title = element_blank())


```



```{r chunk-82}


cyclistic_merged %>%
ggplot(aes(day_of_week, fill = member_casual)) +
geom_bar(position = "dodge") +
facet_wrap(~ month) +
scale_y_continuous( labels = scales::comma) +
labs(title = "Number of Rides",
subtitle = "weekday compared by month",
caption = "Data source: Motivate International Inc. Figure 2",
y = "Number of Rides",
fill = "Rider type") +
xlab("") +
theme(axis.text.x = element_text(angle = 90), legend.title = element_blank())


```


```{r chunk-83}

cyclistic_merged %>%
  group_by(member_casual, month) %>%
  summarise(num_rides = n(), .groups = 'drop') %>%
  ggplot(aes(x = month, y = num_rides, fill = member_casual)) +
  geom_col(position = "dodge", width = .5) +
  scale_y_continuous(labels = scales::comma) +
  labs(
      title = "Number of Rides",
      subtitle = "by Month",
      caption = "Data source: Motivate International Inc. Figure 3",
      x = "",
      y = "Number of Rides") +
theme(axis.text.x = element_text(angle = 90), legend.title = element_blank())





```


```{r chunk-84}


cyclistic_merged %>%
  group_by(member_casual, hour) %>%
  summarise(num_rides = n(), .groups = 'drop') %>%
  ggplot(aes(x = hour, y = num_rides, fill = member_casual), las = 2,
  cex.names = 0.1) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Number of Rides",
    subtitle = "by Hour",
    caption = "Data source: Motivate International Inc. Figure 4",
    x = " ",
    y = "Number of Rides") +
scale_x_continuous(breaks = seq(0,24, by = 6)) +
theme(legend.title = element_blank())


```


```{r chunk-85}


cyclistic_merged %>%
  ggplot(aes(hour, fill = member_casual)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ month) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Number of Rides",
       subtitle = "monthly by hour",
       caption = "Data source: Motivate International Inc. Figure 5",
       y = "Number of Rides",
       fill = "Rider type") +
scale_x_continuous(breaks = seq(0,24, by = 4)) +
xlab("") +
theme(axis.text.x = element_text(angle = 90), legend.title = element_blank())



```


```{r chunk-86}

cyclistic_merged %>% filter(rideable_type != "docked_bike") %>% ## docked bikes are only used by casual
  group_by(member_casual, rideable_type) %>%
  count() %>%
  group_by(member_casual) %>%
  mutate(percentage = n/sum(n) * 100) %>%
  ggplot(aes(x=member_casual, y=percentage, fill=rideable_type)) +
  geom_col(position="dodge") +
  geom_text(aes(label = paste0(round(percentage), "%")),
            position=position_dodge(0.85), vjust = 0) +
labs(x="Rider Type",
     y="Percentage (%)",
     title="Percentage of Rides",
     subtitle = "by bike type",
     caption = "Data source: Motivate International Inc. Figure 6") +
theme(legend.title = element_blank())


```


```{r chunk-87}

cyclistic_merged %>% filter(member_casual == "member") %>%
  group_by(start_station_name) %>%
  summarise(total = n()) %>%
  arrange(desc(total)) %>%
  slice_head(n = 3) %>%
  ggplot(aes(x = reorder(start_station_name, total), y = total,
             fill=start_station_name)) +
  scale_y_continuous(labels = scales::comma) +
  geom_col(width = 0.65) +
  coord_flip() +
  labs(y = "Number of Rides",
       x = "Start Station",
       caption = "Data source: Motivate International Inc. Figure 7",
       title = "3 Most Frequented Start Stations",
       subtitle = "by member user") +
geom_text(aes(label = total), color="white", hjust=1.1) +
theme(legend.title = element_blank())


```


```{r chunk-88}


cyclistic_merged %>% filter(member_casual == "casual") %>%
  group_by(start_station_name) %>%
  summarise(total = n()) %>%
  arrange(desc(total)) %>%
  slice_head(n = 3) %>%
  ggplot(aes(x = reorder(start_station_name, total), y = total,
             fill=start_station_name)) +
  scale_y_continuous(labels = comma) +
  geom_col(width = 0.65) +
  coord_flip() +
  labs(y = "Number of Rides",
       x = "Start Station",
       caption = "Data source: Motivate International Inc. Figure 8",
       title = "3 Most Frequented Start Stations",
       subtitle = "by casual user") +
geom_text(aes(label = total), color="white", hjust=1.1) +
theme(legend.title = element_blank())





```


```{r chunk-89}

cyclistic_merged %>% 
  filter(member_casual == "casual") %>%
  group_by(start_station_name) %>%
  summarise(total = n()) %>%
  arrange(desc(total)) %>%
  slice_head(n = 3) %>%
  ggplot(aes(x = reorder(start_station_name, total), y = total,
             fill=start_station_name)) +
  scale_y_continuous(labels = comma) +
  geom_col(width = 0.65) +
  coord_flip() +
  labs(y = "Number of Rides",
       x = "Start Station",
       caption = "Data source: Motivate International Inc. Figure 9",
       title = "3 Most Frequented Start Stations",
       subtitle = "by casual user") +
geom_text(aes(label = total), color="white", hjust=1) +
theme(legend.title = element_blank())
27


```


```{r chunk-90}

cyclistic_merged %>% 
  filter(member_casual == "member") %>%
  group_by(end_station_name) %>%
  summarise(total = n()) %>%
  arrange(desc(total)) %>%
  slice_head(n = 3) %>%
  ggplot(aes(x = reorder(end_station_name, total), y = total,
  fill=end_station_name)) +
  scale_y_continuous(labels = comma) +
  geom_col(width = 0.65) +
  coord_flip() +
  labs(y = "Number of Rides",
       x = "End Station",
       caption = "Data source: Motivate International Inc. Figure 10",
       title = "3 Most Frequented End Stations",
       subtitle = "by member user") +
geom_text(aes(label = total), color="white", hjust=1) +
theme(legend.title = element_blank())



```

```{r chunk-91}

cyclistic_merged %>% 
  filter(member_casual == "casual") %>%
  group_by(end_station_name) %>%
  summarise(total = n()) %>%
  arrange(desc(total)) %>%
  slice_head(n = 3) %>%
  ggplot(aes(x = reorder(end_station_name, total), y = total,
  fill=end_station_name)) +
  scale_y_continuous(labels = comma) +
  geom_col(width = 0.65) +
  coord_flip() +
  labs(y = "Number of Rides",
       x = "End Station",
       caption = "Data source: Motivate International Inc. Figure 11",
       title = "3 Most Frequented End Stations",
       subtitle = "by casual user") +
geom_text(aes(label = total), color="white", hjust=1) +
theme(legend.title = element_blank())




```

# Findings

#### *	Members account for 55.8% of rides and casual riders account for the remaining 44.2%.
#### *	67.5 % of all rides are taken between Monday and Friday. 
#### *	Member riders are highest Monday through Friday between the hours of 5am and 8 pm.
#### *	Casual riders are highest weeknights between the hours of 9pm and 4am and Saturday and Sunday



# Recommendations
Information was not provided to connect number of rides to a particular rider.


1. Promote "Ride with a Friend" as a membership perk on low ridership days (Monday & Tuesday).
2. Market "Half-Price" riders during the spring in hopes of creating a pre-summer exercise routine.
3. Offer a weekend only plan at a premium. 


